\Large Generative Dynamics of Supreme Court Citations: \\ Analysis with a New Statistical Model

### 2020-05-04 Tex/paper.tex
completed editing pass

\maketitle
\begin{abstract}
\noindent The significance and influence of US Supreme Court majority opinions derive in large part from opinions' roles as precedents for future opinions. A growing body of literature seeks to understand what drives the use of opinions as precedents through the study of Supreme Court case citation patterns. We raise two limitations of existing work on Supreme Court citations. First, dyadic citations are typically aggregated to the case level before they are analyzed. Second, citations are treated as if they arise independent of one another. We make the case that studying Supreme Court citations as a dynamic citation network overcomes both of these limitations. We present a methodology for studying citations between US Supreme Court opinions at the dyadic level, as a network. This methodology---the citation temporal exponential random graph model---enables researchers to account for the effects of case characteristics and complex forms of network dependence in citation formation. We apply this methodology to a network that includes all Supreme Court cases decided between 1950 and 2015. We find evidence for dependence processes, including reciprocity, transitivity, popularity, and activity. The dependence effects that we identify are as substantively and statistically significant as the effects of the exogenous covariates we include in the model. The summary result from this analysis is that theoretical models of Supreme Court citation formation should consider both the effects of case characteristics and the structure of past citations.\footnote{
%Prepared for presentation at the 2018 Annual Meeting of the American Political Science Association. 
This work was supported in part by NIH, 
R01 AI36664-01. NSF grants SES- 1558661, SES-1637089, SES-1619644, and CISE-1320219. We would like to thank Rachael Hinkle and Benjamin Kassow for detailed comments on an earlier draft of this paper.}
\end{abstract}

@@

We include dependence terms to account for each of the dynamics that we hypothesize will characterize the Supreme Court Citation Network---transitivity, reciprocity, popularity, and activity. %We expect that the coefficients associated with each of these dependence effects will be positive.

@@

Our case for studying legal citations at the directed dyadic level hinges upon the contribution to modeling offered by incorporating network dependence. To quantify this contribution, we use AIC and BIC out-of-sample prediction. Predicting out-of-sample offers an unbiased and general purpose way to evaluate the contribution, in terms of model fit, of one or more terms/parameters in a model \citep{jensen2000multiple,ward2010perils}. Unlike in-sample measures of model fit, out-of-sample methods are highly robust in avoiding overfitting, and work when we cannot accurately calculate the value of the likelihood function, as in the current case. Out-of-sample prediction is a common way to evaluate methods for modeling ties in networks, and has recently been applied to TERGMs in particular \citep{desmarais2013forecasting,cranmer2017can}.

@@

Figure \ref{AIC_BIC} depicts the AIC and BIC comparison for two different models for each term from 1950 to 2015. The full model is defined as the model that includes all covariate terms and dependence terms described in \ref{covariate_terms} and \ref{dependence_terms}. The independent model only consists of covariate terms, not accounting for any structural dependencies in the networks. We see that the AIC and BIC is considerably smaller for the full model throughout the period of consideration. On average the independent model yields a 5525 point higher AIC and an average 5582 point higher BIC. The maximum difference is reached in 1983 with a difference of 10,990 points for the AIC and 10,930 points for the BIC.

@@

\subsection{c-TERGM Results}
The results from the c-TERGM are presented Figures \ref{SCC_results_1} and \ref{SCC_results_2}. In these figures we depict coefficient estimates and 95\% confidence intervals for all of the effects for all the variables we discussed in sections \ref{covariate_terms} and \ref{dependence_terms}. The coefficients in an ERGM family model can be interpreted in the same way as logistic regression coefficients---they give the change in the log odds of a tie from $i$ to $j$ given a one unit increase in the respective variable. For this reason, the number of edges in a network corresponds to the intercept in an ERGM. We first discuss the dependence effect estimates in Figure \ref{SCC_results_1}. A circle indicates that the estimate was significant at an $\alpha=0.05$ level, while and square translates to a p-value between $0.05$ and $0.1$. A triangle signals that the estimate for a given year was not significant at an $\alpha=0.1$ level.

The panel showing the results of the reciprocity statistic does not provide estimates for ever term. In these missing terms there were no two cases that would cite each other, making the estimation of the reciprocity statistic impossible. A similar scenario is visible for the \textit{Overruled Cases} variable. In the time period from 1953-1956 no cases that have been overruled have been cited which results in a non-estimable effect.

%The reciprocity effect is substantial, positive, and statistically significant for most of the period under study. Conditional upon cases $i$ and $j$ being considered within the same term, the log of the odds that opinion $i$ cites opinion $j$ increases by approximately $2$ when opinion $j$ cites opinion $i$. The naive probability that any one case cites another case decided at the same time or previously in the Supreme Court Citation network is approximately $0.0019$. This probability increases eight-fold, to approximately $0.015$ if the citation would be reciprocal. The transitivity (i.e., triangle closure) effect is also substantial, positive, and statistically significant over the entire period studied. The probability that case $i$ cites case $j$ increases approximately five-fold if the citation would close a triangle involving cases $i$ and $j$. The activity and popularity effects are both estimated to be positive, but trend downward and are more modest in scale (and in the case of activity, not statistically significant later in the time series). Interpreting the popularity effect---the log odds that $i$ cites $j$ increases by 0.018--0.035 with every additional citation to case $j$. Based on the indegree distribution, it appears that the popularity of cases can be classified by multiples of 10. If case $j$ increases from ten to twenty citations from other cases, the probability that case $i$ cites case $j$ increases from approximately 0.0030 to 0.0040. Overall, these results represent evidence of complex dependence in Supreme Court citation formation.

%The effects of the exogenous covariates included in the c-TERGM (panels (e)--(k) in Figure \ref{fig:coeftrends_exo}) exhibit some surprising dynamics. The effects of absolute difference in Martin-Quinn score, and majority size of the recipient case all flip signs throughout the time period under study. This may be an artifact of the simple linear trends we have estimated, but is interesting nonetheless. The attitudinal model holds that ideological preferences drive decision-making on the US Supreme Court. Ideological selection of precedents would be evidenced by a negative effect of the absolute difference in MQ scores in the c-TERGM. We find this effect is negative and statistically significant only after 1975. The absolute difference in MQ scores variable is on the scale of approximately 0---3. At its most ideological point (i.e. 2001), the probability that case $i$ cites case $j$ decreases from approximately 0.024 to 0.018---by one third---if the ideological distance between the majority coalitions in the two cases increases by 3. The effects of the majority size and ideological breadth of case $j$ are similar in both scale and trend. It is not until later in the time series that they both exhibit their expected positive effects. The effect of same issue is positive and statistically significant, and substantial for the time period under study. The probability that $i$ cites $j$ increases by approximately three-fold over the naive probability of citation if $i$ and $j$ have the same legal issue area.

@@

%{\em {\bf Note:} The results section of this paper is incomplete and in flux. We recently discovered evidence of non-convergence in the C-TERGM results currently presented in the paper, and have changed the models and estimation settings to address this problem. Most of the substantive conclusions look like they will hold using the updated models, but we are not sure that we will be able to include the updated models in the APSA presentation.}

@@

We present a methodology for studying citations between US Supreme Court opinions at the dyadic level, as a network. This methodology---the citation-ERGM---enables researchers to include both exogenous covariates such as the ideological predisposition and age of a case, and dependence terms, such as transitivity and reciprocity, as explanations for citation formation. We apply this methodology to a network that includes all Supreme Court cases decided between 1937 and 2015. We find, somewhat counterintuitively, that Supreme Court citations are highly reciprocal. We also find that citations are driven by dependencies such as triad closure, popularity, and activity. The dependence effects that we identify are as substantively and statistically significant as the effects of the exogenous covariates we include in the model. The summary result from this analysis is that theoretical models of Supreme Court citation formation should consider both the effects of case characteristics and the structure of past citations.

We recognize several limitations of the current analysis, which should be addressed in future iterations of this work or in future studies. First, our theoretical claims and analyses treat all citations---positive and negative---as equivalent. Both data and methodological limitations will need to be overcome to consider these two citation types. Second, there is nothing in the c-ERGM that governs the generation of new cases (i.e., cases are assumed to arise exogenously, unrelated to the history of the citation network). As new cases are often taken up by the Supreme Court in order to test, clarify, and/or challenge existing precedents, it is likely inappropriate to assume that new cases arise exogenous to the existing citation network. Overcoming this limitation would require methodological innovation in the c-ERGM as well as theoretical development regarding exactly how citation structure might predict case emergence. 
%Third, the data used in this study does not cover the most recent history of the Court. Our data ends in 2001---missing the tenure of the most recent chief justice, and more than fifteen terms. To overcome this limitation we would need to extend the citation data beyond that collected by Fowler and colleagues.

>>>>

\maketitle
\begin{abstract}
\noindent The significance and influence of US Supreme Court majority opinions derive in large part from opinions' roles as precedents for future opinions. A growing body of literature seeks to understand what drives the use of opinions as precedents through the study of Supreme Court case citation patterns. We raise two limitations of existing work on Supreme Court citations. First, dyadic citations are typically aggregated to the case level before they are analyzed. Second, citations are treated as if they arise independent of one another. We make the case that studying Supreme Court citations as a dynamic citation network overcomes both of these limitations. We present a methodology for studying citations between US Supreme Court opinions at the dyadic level, as a network. This methodology---the citation temporal exponential random graph model---enables researchers to account for the effects of case characteristics and complex forms of network dependence in citation formation. We apply this methodology to a network that includes all Supreme Court cases decided between 1950 and 2015. We find evidence for dependence processes, including reciprocity, transitivity, and popularity. The dependence effects that we identify are as substantively and statistically significant as the effects of the exogenous covariates we include in the model. The summary result from this analysis is that theoretical models of Supreme Court citation formation should consider both the effects of case characteristics and the structure of past citations.\footnote{
%Prepared for presentation at the 2018 Annual Meeting of the American Political Science Association. 
This work was supported in part by NIH, 
R01 AI36664-01. NSF grants SES- 1558661, SES-1637089, SES-1619644, and CISE-1320219. We would like to thank Rachael Hinkle and Benjamin Kassow for detailed comments on an earlier draft of this paper.}
\end{abstract}

@@

We include dependence terms to account for each of the dynamics that we hypothesize will characterize the Supreme Court Citation Network---transitivity, reciprocity, and popularity. %We expect that the coefficients associated with each of these dependence effects will be positive.

@@

Our case for studying legal citations at the directed dyadic level hinges upon the contribution to modeling offered by incorporating network dependence. To quantify this contribution, we fit one model that incorporates covariate effects only, which we term the {\em Independent Model}---excluding the reciprocity, transitivity, and popularity terms. The full model is the one we present below. We use both AIC and BIC to compare the fit of the two models, with lower values indicating better fit \citep{ripplinger2008does}.

@@

Figure \ref{AIC_BIC} depicts the AIC and BIC comparison for the two different models for each term from 1950 to 2015. We see that the AIC and BIC is considerably smaller for the full model throughout the period of consideration. On average the independent model yields a 5,525 point higher AIC and an average 5,582 point higher BIC. The maximum difference is reached in 1983 with a difference of 10,990 points for the AIC and 10,930 points for the BIC. This fit comparison provides robust evidence that the dependence terms contribute significantly to the explanatory power of the model.\footnote{In the appendix we present conventional goodness-of-fit plots that are used to assess the structural fit for ERGM \citep{hunter2008goodness}, as well as degeneracy diagnostics \citep{mukherjee2020degeneracy}. We find that the full model fits the degree and edgewise shared partner distributions well, and is not degenerate.}

@@

\subsection{c-ERGM Coefficient Estimates}
The estimates from the c-ERGM are presented in Figures \ref{SCC_results_1} and \ref{SCC_results_2}. In these figures we depict coefficient estimates and 95\% confidence intervals for all of the effects for all the variables we discussed in sections \ref{covariate_terms} and \ref{dependence_terms}. The coefficients in an ERGM family model can be interpreted in the same way as logistic regression coefficients---they give the change in the log odds of a tie from $i$ to $j$ given a one-unit increase in the respective variable. We first discuss the dependence effect estimates in Figure \ref{SCC_results_1}. A circle indicates that the estimate was significant at an $\alpha=0.05$ level, while and square translates to a p-value between $0.05$ and $0.1$. A triangle signals that the estimate for a given year was not significant at an $\alpha=0.1$ level. The panel showing the results of the reciprocity statistic does not provide estimates for every term. In these missing terms there were no two cases that cited each other, making the estimate of the reciprocity statistic term equivalent to $-\infty$. For those years, the reciprocity term is included in the model, but the estimate of $-\infty$. A similar scenario is visible for the \textit{Overruled Cases} variable. In the time period from 1953-1956 no cases that have been overruled have been cited which results in a non-estimable effect.

We first discuss our findings regarding the dependence terms. In most of the years that it is estimable, the reciprocity effect is statistically significantly positive, and substantial in magnitude. In a typical year, the presence of a citation from case $i$ to case $j$ increases the log-odds of a citation from case $j$ to case $i$ by approximately 0.50. Both transitivity effects---different term transitivity, and GWESP---are estimated to be positive and statistically significant in every year, and are even greater in magnitude than the reciprocity effect. The log-odds of a citation from case $i$ to case $j$ increases by more than 2 if there is at least one third case, $k$ that is cited by $i$, and cites to $j$. Results regarding the popularity effect (i.e., GWIdegree) are generally supportive of the hypothesis of a popularity effect in Supreme Court citations, as the GWIdegree estimate is negative in every term. There are some terms for which the result is not statistically significant at conventional levels. The popularity effect looks to have grown substantially stronger after the 2000 term. Contrary to expectations, we find that the hub effect is negative. That is, each additional citation sent by a case reduces the likelihood that the case is cited. For every citation sent, the log-odds of any citation to a case decreases by approximately 0.025---a result that is statistically significant in each term.

The effects of the exogenous covariates included in the c-ERGM are presented in Figure \ref{SCC_results_2}.  Most of the exogenous covariate effects align with expectations. We find that cases are more likely to cite those that (1) have been decided most recently, (2) were decided by large majorities, (3) have not been overruled, (3) were authored by the same justice, and (3) are in the same issue area. Surprisingly, we do not find consistent effects of either of the covariates that involve justice ideology. The ideological distance between two cases, as measured by the absolute difference in the median of the majority coalitions; and the ideological breadth of the cited cases, as measured by the absolute difference between the most liberal and most conservative justices on the potential receiving case, both exhibit signs and significance levels that are highly volatile. Our results leave open the question of whether, and if so, how, citations are shaped by ideological factors.

@@

@@

We present a methodology for studying citations between US Supreme Court opinions at the dyadic level, as a network. This methodology---the citation-ERGM---enables researchers to include both exogenous covariates such as the ideological predisposition and age of a case, and dependence terms, such as transitivity and reciprocity, as explanations for citation formation. We apply this methodology to a network that includes all Supreme Court cases decided between 1950 and 2015. We find, somewhat counterintuitively, that Supreme Court citations are highly reciprocal. We also find that citations are driven by dependencies such as triad closure and popularity. The dependence effects that we identify are as substantively and statistically significant as the effects of the exogenous covariates we include in the model. The summary result from this analysis is that theoretical models of Supreme Court citation formation should consider both the effects of case characteristics and the structure of past citations.

We recognize two important limitations of the current analysis, which should be addressed in future iterations of this work or in future studies. First, our theoretical claims and analyses treat all citations---positive and negative---as equivalent. Both data and methodological limitations will need to be overcome to consider these two citation types. Second, there is nothing in the c-ERGM that governs the generation of new cases (i.e., cases are assumed to arise exogenously, unrelated to the history of the citation network). As new cases are often taken up by the Supreme Court in order to test, clarify, and/or challenge existing precedents, it is likely inappropriate to assume that new cases arise exogenous to the existing citation network. Overcoming this limitation would require methodological innovation in the c-ERGM as well as theoretical development regarding exactly how citation structure might predict case emergence. 
%Third, the data used in this study does not cover the most recent history of the Court. Our data ends in 2001---missing the tenure of the most recent chief justice, and more than fifteen terms. To overcome this limitation we would need to extend the citation data beyond that collected by Fowler and colleagues.
